---
title: "BulkTranscrptomicsSept2026"
output: html_document
date: "2026-09-27"
---

##################################
Loading Libraries for Bulk RNASeq#
##################################
```{r pressure0, echo=FALSE}

library(conflicted)
library(stringr)
library(dplyr)
library(plyr)
library(tidyr)
library(tidyverse)
library(tibble)
library(ggpubr)
library(rstatix)
library(ggprism)
library(patchwork)
library(magrittr)
library(data.table) 
library(fuzzyjoin)
library(janitor)


#Analyses
library(DESeq2)
library(topGO)
library (clusterProfiler)
library(R.utils)
library(enrichplot)
library(GO.db)
library(variancePartition)


#Figs
library(ggplot2)
library(ggrepel)
library(ggsignif)
library(ComplexHeatmap)
library(cowplot)
library(circlize)
library(RColorBrewer)
library(EnhancedVolcano)
```





###########################################
Loading in Data and Setting up DESeq Table#
###########################################
```{r}
setwd("/Users/johnbriseno/Desktop/Manuscripts/Ch2_Bulk_Atlas/Supplemental/Bulk Supp./")
```

```{r pressure1, echo=FALSE}

###########################
###Inputting Count Tables##
###########################

counts_table <- file.choose() 
#look for featureCounts.matrix.txt

featureCounts <- read.table(counts_table,
  header = TRUE,
  row.names=1
  )

#testing to get names of columns so i can rename them proper sample names
#removing rly long path string from col headers put there by feature counts
colnames(featureCounts) <- gsub("X.home.FCAM.jbriseno.JB_Nyholm_Lab.Sept2026_Ch2_BulkDataV3_ReAnalyses.STAR_GenomeV3_BRAKER_BULK.BAM..", "", colnames(featureCounts))

#making samples dict/key to quickly rename
#https://stackoverflow.com/questions/59314285/selectively-rename-r-data-frame-column-names-using-a-key-value-pair-dictionary
idkey <- data.frame("fc_col_order" =  c("167717_S2Aligned.sortedByCoord.out.bam",
                                        "167719_S4Aligned.sortedByCoord.out.bam",
                                        "167720_S5Aligned.sortedByCoord.out.bam",
                                        "167721_S6Aligned.sortedByCoord.out.bam",
                                        "167722_S7Aligned.sortedByCoord.out.bam",
                                        "167728_S13Aligned.sortedByCoord.out.bam",
                                        "167730_S15Aligned.sortedByCoord.out.bam",
                                        "167734_S19Aligned.sortedByCoord.out.bam",
                                        "167736_S21Aligned.sortedByCoord.out.bam",
                                        "167737_S22Aligned.sortedByCoord.out.bam",
                                        "167741_S26Aligned.sortedByCoord.out.bam",
                                        "167743_S28Aligned.sortedByCoord.out.bam",
                                        "167744_S29Aligned.sortedByCoord.out.bam",
                                        "167745_S30Aligned.sortedByCoord.out.bam",
                                        "167747_S32Aligned.sortedByCoord.out.bam",
                                        "167749_S34Aligned.sortedByCoord.out.bam",
                                        "167751_S36Aligned.sortedByCoord.out.bam",
                                        "190853_S1Aligned.sortedByCoord.out.bam",
                                        "190854_S2Aligned.sortedByCoord.out.bam",
                                        "190904_S3Aligned.sortedByCoord.out.bam",
                                        "190905_S4Aligned.sortedByCoord.out.bam",
                                        "190906_S5Aligned.sortedByCoord.out.bam",
                                        "190907_S6Aligned.sortedByCoord.out.bam",
                                        "190908_S7Aligned.sortedByCoord.out.bam",
                                        "190909_S8Aligned.sortedByCoord.out.bam",
                                        "190910_S9Aligned.sortedByCoord.out.bam",
                                        "190911_S10Aligned.sortedByCoord.out.bam",
                                        "190912_S11Aligned.sortedByCoord.out.bam",
                                        "190913_S12Aligned.sortedByCoord.out.bam",
                                        "190914_S13Aligned.sortedByCoord.out.bam",
                                        "190915_S14Aligned.sortedByCoord.out.bam",
                                        "190916_S15Aligned.sortedByCoord.out.bam",
                                        "190917_S16Aligned.sortedByCoord.out.bam",
                                        "190918_S17Aligned.sortedByCoord.out.bam",
                                        "190920_S19Aligned.sortedByCoord.out.bam",
                                        "190921_S20Aligned.sortedByCoord.out.bam",
                                        "190922_S21Aligned.sortedByCoord.out.bam",
                                        "190923_S22Aligned.sortedByCoord.out.bam",
                                        "190924_S23Aligned.sortedByCoord.out.bam",
                                        "190925_S24Aligned.sortedByCoord.out.bam",
                                        "190931_S28Aligned.sortedByCoord.out.bam",
                                        "190932_S29Aligned.sortedByCoord.out.bam",
                                        "190933_S30Aligned.sortedByCoord.out.bam",
                                        "190934_S31Aligned.sortedByCoord.out.bam",
                                        "190935_S32Aligned.sortedByCoord.out.bam",
                                        "190936_S33Aligned.sortedByCoord.out.bam",
                                        "190938_S35Aligned.sortedByCoord.out.bam",
                                        "190939_S36Aligned.sortedByCoord.out.bam",
                                        "190951_S48Aligned.sortedByCoord.out.bam",
                                        "190953_S50Aligned.sortedByCoord.out.bam",
                                        "190954_S51Aligned.sortedByCoord.out.bam",
                                        "190965_S53Aligned.sortedByCoord.out.bam"
                                        ), 
                    "ideal_name" = c("MS3_MNTL_167717",
                                     "MS3_ARM_167719",
                                     "MS3_SKIN_167720",
                                     "MS3_TST_167721",
                                     "MS3_OLL_167722",
                                     "MS4_OLL_167728",
                                     "MS4_WBL_167730",
                                     "MS4_TST_167734",
                                     "FS13_OLL_167736",
                                     "FS13_WBL_167737",
                                     "FS13_OVRY_167741",
                                     "FS13_LO_CC_167743",
                                     "FS15_CB_167744",
                                     "FS15_OLL_167745",
                                     "FS15_WBL_167747",
                                     "FS13_OVRY_167749",
                                     "FS15_LO_CC_167751",
                                     "MS4_MNTL_190853",
                                     "MT8_ARM_190854",
                                     "MT8_SKIN_190904",
                                     "MT8_TST_190905",
                                     "MT8_OLL_190906",
                                     "MT8_WBL_190907",
                                     "MT8_LO_CC_190908",
                                     "MT8_CB_190909",
                                     "MT13_MNTL_190910",
                                     "MT13_ARM_190911",
                                     "MT13_SKIN_190912",
                                     "MT13_WBL_190913",
                                     "MT13_CB_190914",
                                     "MT14_TST_190915",
                                     "MT14_WBL_190916",
                                     "MT14_LO_CC_190917",
                                     "MT20_WBL_190918",
                                     "MT18_LO_CC_190920",
                                     "MT6_CB_190921",
                                     "FS11_MNTL_190922",
                                     "FS11_ARM_190923",
                                     "FS11_SKIN_190924",
                                     "FS11_WBL_190925",
                                     "FU3_MNTL_190931",
                                     "FU3_ARM_190932",
                                     "FU3_SKIN_190933",
                                     "FU3_OVRY_190934",
                                     "FU3_OLL_190935",
                                     "FU3_WBL_190936",
                                     "FU3_LO_CC_190938",
                                     "FU3_CB_190939",
                                     "FT21_WBL_190951",
                                     "FT21_LO_CC_190953",
                                     "FT12_CB_190954",
                                     "FS9_OVRY_190965"))
colnames(featureCounts) <- dplyr::recode(
  colnames(featureCounts), 
  !!!setNames(as.character(idkey$ideal_name), idkey$fc_col_order)
)




#setting up samples
samples <- c(
  #♂ WBLs
  "MS4_WBL_167730", 
  "MT13_WBL_190913", 
  "MT14_WBL_190916", 
  "MT20_WBL_190918", 
  "MT8_WBL_190907",
  #♀ WBLs
  "FS11_WBL_190925", 
  "FS13_WBL_167737", 
  "FS15_WBL_167747", 
  "FT21_WBL_190951", 
  "FU3_WBL_190936", 
  
  #♂ CBs
  "MT8_CB_190909", 
  "MT13_CB_190914", 
  "MT6_CB_190921",
  #♀ CBs
  "FS15_CB_167744", 
  "FU3_CB_190939", 
  "FT12_CB_190954",
  
  #♂ OLLs
  "MS3_OLL_167722", 
  "MS4_OLL_167728", 
  "MT8_OLL_190906", 
  #♀ OLLs
  "FS13_OLL_167736", 
  "FS15_OLL_167745", 
  "FU3_OLL_190935", 
  
  #♂LO_CC
  "MT8_LO_CC_190908",
  "MT14_LO_CC_190917",
  "MT18_LO_CC_190920",
  #♀ LO_CC
  "FS13_LO_CC_167743",
  "FS15_LO_CC_167751",
  "FU3_LO_CC_190938",
  "FT21_LO_CC_190953",
  
  #OVRYs
  "FS13_OVRY_167749",
  "FS15_OVRY_167741",
  "FS9_OVRY_190934",
  "FU3_OVRY_190965",
  #TSTs
  "MS3_TST_167721",
  "MS4_TST_167734",
  "MT14_TST_190905",
  "MT8_TST_190915",
  
  #♂ MNTLs
  "MS3_MNTL_167717",
  "MS4_MNTL_190853",
  "MT13_MNTL_190910",
  #♀ MNTLs
  "FS11_MNTL_190922",
  "FU3_MNTL_190931",
  #♂ SKIN
  "MS3_SKIN_167720",
  "MT8_SKIN_190904",
  "MT13_SKIN_190912",
  #♀ SKIN
  "FS11_SKIN_190924",
  "FSU3_SKIN_190933",
  #♂ HECTOARM
  "MS3_ARM_167719",
  "MT8_ARM_190854",
  "MT13_ARM_190911",
  #♀ HECTO_ARM
  "FS11_ARM_190923",
  "FU3_ARM_190932"
)

#Need to reorder the counts file so DESeq2 can analyze properly! 
#This gets rid of featureCounts command headers, amazing! 
#https://www.jianshu.com/p/80c7bf8a2834
#colnames(featureCounts) <- c("Chr", "Start", "End", "Strand", "Length", samples)
col_order <- samples
featureCounts_justCOUNTS <- featureCounts[ , col_order]
```





#########################################################
Adding NCBI functional annotation info to counts table###
#########################################################
```{r pressure2, echo=FALSE}

######################################
#Setting gene names for counts matrix#
######################################

#look for GCF_053919585.1_ASM5391958v1_feature_table.txt
ncbi_path <- file.choose()

#reading in file with specific columns:
#LOC gene ID
#descriptive name
#NCBI geneID (used for GO annot I think?)
ncbi_features <- as.data.frame(fread(ncbi_path, select = c("symbol", "name" #"GeneID"
                                                           )))

##remove duplicated rows 
#for 'uncharacterized' transcript variants/isoforms these rows are duplicate
#https://stackoverflow.com/questions/13967063/remove-duplicated-rows
unique_ncbi_features <- unique(ncbi_features[ , 1:2 ] )

#remove duplicated genes (i.e. transcript variants/isoforms)
#just want GENES!
clean_ncbi_features <- unique_ncbi_features[!duplicated(unique_ncbi_features$symbol), ]
#gene count is 39,658

#drop duplicate string LOC names in 'name' column
clean_ncbi_features$name <- gsub("LOC.*", "", clean_ncbi_features$name)

#merge LOC and name together in new column
clean_ncbi_features$Description <- paste(clean_ncbi_features$symbol,
                               clean_ncbi_features$name)

#change spaces and other symbols to underscores in new Description column
#https://stackoverflow.com/questions/53107014/replace-space-between-two-words-with-an-underscore-in-a-vector
clean_ncbi_features$Description <- gsub(" ", "_",
                                       clean_ncbi_features$Description)
#will start with spaces and see if i need to delete anything else down the road 

#merging new names with counts table
#first make rownames new column named symbol in counts table
featureCounts_symbol <- tibble::rownames_to_column(featureCounts_justCOUNTS, "symbol")
merged_df <- merge(clean_ncbi_features, featureCounts_symbol, by = "symbol", all.x=TRUE, all.y=TRUE)

#drop NA (remember these are tRNAs where gene symbol is like 'Trnar-ucg')
#https://stackoverflow.com/questions/4862178/remove-rows-with-all-or-some-nas-missing-values-in-data-frame
merged_df <- na.omit(merged_df)
#gene count is 29,352

#renaming RUNX1 and HHEX since they are 'uncharacterized' in NCBI annots
#also gonna rename PGRP5 and HEPHL1
merged_df$Description <-revalue(merged_df$Description,
                                   c("LOC145291535_uncharacterized_"="LOC145291535_RUNX1*", #RUNX1
                                     "LOC145267028_uncharacterized_"="LOC145267028_HHEX*", #HHEX
                                     "LOC145291335_peptidoglycan-recognition_protein_SC2-like"="LOC145291335_EsPGRP5" #PGRP5
                                     ))


#save merged df cause will probably have to change gene names back to symbol for EnTAP GO analyses
write.csv(merged_df, "NCBI_LOC_Descript_counts_matrix.csv")
```

- When needing to work with merged_df again, just read it in as csv instead of running thru above block
```{r}
#merged_df <- read.csv("NCBI_LOC_Descript_counts_matrix.csv")

#dropping other columns
drops <- c("symbol","name")
final_counts <- merged_df[ , !(names(merged_df) %in% drops)]

#resetting rownames with new names
rownames(final_counts) <- final_counts$Description
final_counts <- subset(final_counts, select = -c(Description))

#find duplicated rows
final_counts[duplicated(final_counts$Description),]

#save final counts
saveRDS(final_counts, file = "final_counts.rds")
# Write to CSV
write.csv(final_counts, file = "final_counts.csv", row.names = TRUE)
```

Reading in final_counts
```{r}
# Read back in
final_counts <- readRDS("final_counts.rds")
final_counts <- read.csv("final_counts.csv", row.names = 1)
```


#############################
Generating a samples table###
#############################
```{r pressure3, echo=FALSE}

Tissue.Type <- factor(c("M.WB", "M.WB", "M.WB", "M.WB", "M.WB", 
                         "F.WB", "F.WB", "F.WB", "F.WB", "F.WB",
                         
                         #♀/♂ CBs
                         "M.CB", "M.CB", "M.CB",
                         "F.CB", "F.CB", "F.CB",
                         
                         #♀/♂ OLs
                         "M.OL", "M.OL", "M.OL",
                         "F.OL", "F.OL", "F.OL",
                         
                         #♀/♂ LO_CCs
                         "M.LO_CC", "M.LO_CC", "M.LO_CC",
                         "F.LO_CC", "F.LO_CC", "F.LO_CC", "F.LO_CC",
                         
                         
                         #OVRY/TSTs
                         "OVRY", "OVRY", "OVRY", "OVRY",
                         "TST", "TST", "TST", "TST",
                        
                        
                        
                        #♀/♂ MNTLs,
                        "M.MNTL", "M.MNTL", "M.MNTL", 
                        "F.MNTL", "F.MNTL",
                        
                        #♀/♂ SKINs,
                        "M.SKIN", "M.SKIN", "M.SKIN", 
                        "F.SKIN", "F.SKIN",
                        
                        #♀/♂ _ARMs,
                        "M.ARM", "M.ARM", "M.ARM", 
                        "F.ARM", "F.ARM"
                         
))

Batch <- factor(c(#♀/♂ WBLs
                   "1", "2", "2", "2", "2", 
                   "2", "1","1", "2", "2",
                   
                   #♀/♂ CBs
                   "1", "2", "2", 
                   "2", "2", "2",
                   
                   
                   #♀/♂ OLLs
                   "1", "1", "2", 
                   "1", "1", "2",
                   
                   
                   #♀/♂ LO_CCs
                   "2", "2", "2",  
                   "1", "1", "2", "2",
                   
                   
                   #♀ OVRYs
                   "1", "1", "2", "2",
                   #♂ TSTs
                   "1", "1", "2", "2",
                   
                   #♀/♂ MNTLs,
                   "1", "2", "2",
                   "2", "2",
                   
                   #♀/♂ SKINs,
                   "1", "2", "2",
                   "2", "2",
                   
                   #♀/♂ HECTO_ARMs,
                   "1", "2", "2",
                   "2", "2"
                   
                   
                   
                   ))

Organ.Type <- factor(c(
  #♀/♂ WBLs
  "Hematopoietic", "Hematopoietic", "Hematopoietic", "Hematopoietic", "Hematopoietic",
  "Hematopoietic", "Hematopoietic", "Hematopoietic", "Hematopoietic", "Hematopoietic",
  
  #♀/♂ CBs
  "Neural", "Neural", "Neural",
  "Neural", "Neural", "Neural",
  
  #♀/♂ OLLs
  "Neural", "Neural", "Neural",
  "Neural", "Neural", "Neural",
  
  #♀/♂ LO_CCs
  "Symbiotic", "Symbiotic", "Symbiotic",
  "Symbiotic", "Symbiotic", "Symbiotic", "Symbiotic",
  
  #OVRY/TSTs
  "Gonadal", "Gonadal", "Gonadal", "Gonadal",
  "Gonadal", "Gonadal", "Gonadal", "Gonadal",
  
  #MNTLs
  "Epidermal", "Epidermal", "Epidermal",
  "Epidermal", "Epidermal",
  
  #SKINs
  "Epidermal", "Epidermal", "Epidermal",
  "Epidermal", "Epidermal",
  
  #HECTO_ARM
  "Epidermal", "Epidermal", "Epidermal",
  "Epidermal", "Epidermal"
  
  
))

Sex <- factor(c(
  #♀/♂ WBLs
  "M", "M", "M", "M", "M",
  "F", "F", "F", "F", "F",
  
  #♀/♂ CBs
  "M", "M", "M",
  "F", "F", "F",
  
  #♀/♂ OLLs
  "M", "M", "M",
  "F", "F", "F",
  
  #♀/♂ LO_CCs
  "M", "M", "M",
  "F", "F", "F", "F",
  
  #♀/♂ OVRYs/TSTs
  "F", "F", "F", "F",
  "M", "M", "M", "M",


  #♀ APO WBs
  #"F", "F", "F",
  
  #♀/♂ MNTLs
  "M", "M", "M",
  "F", "F",

  #♀/♂ SKIN
  "M", "M", "M",
  "F", "F",
  
  #♀/♂ HECTO_ARM
  "M", "M", "M",
  "F", "F"
  
))


Hemato <- c("Hematopoietic",
            "Hematopoietic",
            "Hematopoietic",
            "Hematopoietic",
            "Hematopoietic",
            "Hematopoietic",
            "Hematopoietic",
            "Hematopoietic",
            "Hematopoietic",
            "Hematopoietic",
            
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic",
            "NonHematopoietic"
            
            
            )

#A contextual data frame
All.coldata <- data.frame(row.names=colnames(final_counts), 
                          Batch, 
                          Sex,
                          Tissue.Type,
                          Organ.Type,
                          Hemato
                          ) 

#save All coldata
saveRDS(All.coldata, file = "All.coldata.rds")
# Write to CSV
write.csv(All.coldata, file = "All.coldata.csv", row.names = TRUE)
```

Read in All.coldata
```{r}
All.coldata <- readRDS("All.coldata.rds")
All.coldata <- read.csv("All.coldata.csv", row.names = 1)
```



###################################
PCA Plots & Data Transformations###
###################################
- rlog-transform or variance-stabilize count values
- Use PCAs to see diffs among samples before and after Seq.Batch effects
- Used "Organ.Type" attribute to generate PCA so have to generate a dds still for M/F WB "tissue.type" comparison (which we do below)
  - See GO/KO analysis below - analyzed WB DEGs from pairwise comparisons of hemato (wbs) to other organ types

9.27 - I completely forgot to order these properly for feature counts!
solutions: rename all the samples by hand to get formatting right from previous script...DONE!

```{r pressure4, echo=FALSE}

#making DESeq object
dds <- DESeqDataSetFromMatrix( 
  countData = final_counts, 
  colData = All.coldata, 
  design = ~ Batch + 
    Sex + 
    Organ.Type 
) 

design(dds)
colData(dds)

#Prefilter
smallestGroupSize <- 7 #7 -> smallest group size for symbiotic organ.type
keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
dds <- dds[keep,]

#DESeq algorithm
dds <- DESeq(dds)

#transformation
vsd <- vst(dds, 
           blind = T)
```

```{r}
#examine pre-batch correct via PCA
plotPCA(vsd,intgroup=c("Batch"))
plotPCA(vsd,intgroup=c("Organ.Type"))

#sequencing batch correction
mat.vsd <- assay(vsd)
mm.vsd <- model.matrix(~Organ.Type, colData(vsd))
mat.vsd <- limma::removeBatchEffect(mat.vsd, 
                                    batch=vsd$Batch,  
                                    design=mm.vsd)
assay(vsd) <- mat.vsd

#examine post-batch correct via PCA
plotPCA(vsd,intgroup=c("Batch"))
plotPCA(vsd,intgroup=c("Organ.Type"))
```

```{r}
#Fig. 3A - PCA code


pca.data <- plotPCA(vsd, 
                    intgroup=c("Organ.Type", "Sex"), 
                    returnData=TRUE,
                    ntop = 500,
                    pcsToUse = 1:2:3
                    
                    )

#solving pca variance (axes)
percentVar <- round(100 * attr(pca.data, "percentVar"))

#change color to what I want
scale_custom <- scale_fill_manual(values = c("#E18004", 
                                             "#0072B2",
                                             "black", 
                                             "yellow4", 
                                             "#009E73"))
ggplot(pca.data, 
       aes(PC1, PC2, 
           fill = Organ.Type, 
           #color = Tissue.Type, 
           shape = Sex, 
           group = Organ.Type
       )) +
    geom_point(aes(PC1,
                 PC2,
                 fill = Organ.Type),
             #color = "black",
             size = 4,
             ) +
  scale_shape_manual(values = c(21, 24)) +
  
  #colors for organ type
  scale_fill_manual(values = c("#E18004", 
                               "#0072B2",
                               "navajowhite", 
                               "yellow4", 
                               "#009E73")) + #your colors here
  
  #axes/labels
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) +
  guides(fill = guide_legend(override.aes = list(shape = 23))) +
  stat_ellipse(data = pca.data,
               aes(color = Organ.Type)
               )
```



#########
3D PCA###
#########
Mixed 3D PCA function from these two sources:
Modified plotPCA from DESeq2 package. Shows the Names of the Samples (the first col of SampleTable), and uses ggrepel pkg to plot them conveniently <- https://www.biostars.org/p/333436/
plot3DPCA from btools <- https://github.com/twbattaglia/btools/blob/master/R/plotPCA3D.R
```{r pressure5, echo=FALSE}
plotPCA3D <- function (object, intgroup = "condition", ntop = 500, returnData = FALSE)
{
  rv <- genefilter::rowVars(assay(object))
  select <- order(rv, decreasing = TRUE)[seq_len(min(ntop, length(rv)))]
  pca <- prcomp(t(assay(object)[select, ])) #PCA function
  percentVar <- pca$sdev^2/sum(pca$sdev^2) #caluclating PCA variance
  if (!all(intgroup %in% names(colData(object)))) {
    stop("the argument 'intgroup' should specify columns of colData(dds)")
  }
  intgroup.df <- as.data.frame(colData(object)[, intgroup, drop = FALSE])
  group <- if (length(intgroup) > 1) {
    factor(apply(intgroup.df, 1, paste, collapse = " : "))
  }
  else {
    colData(object)[[intgroup]]
  }
  ## Select the PCAs and percentVar that you like instead of 1 and 2
  d <- data.frame(PC1 = pca$x[, 1],
                  PC2 = pca$x[, 2],
                  PC3 = pca$x[, 3],
                  group = group,
                  intgroup.df,
                  name = colnames(object))
  if (returnData) {
    attr(d, "percentVar") <- percentVar[1:3]
    return(d)
  }
  message("Generating plotly plot")
  p <- plotly::plot_ly(data = d,
                       x = ~PC1,
                       y = ~PC2,
                       z = ~PC3,
                       color = Organ.Type,
                       #symbol = Sex,
                       symbols = Sex,
                       mode = "markers",
                       type = "scatter3d")
  return(p)
}

plotly_obj <- plotPCA3D(vsd,
                        intgroup = c("Organ.Type",
                                     "Sex"),
                        ntop = 500,
                        returnData = FALSE)
plotly_obj


#changing symbols & colors for Sex & organ type
#https://stackoverflow.com/questions/72555895/how-to-manually-add-symbols-with-plotly-in-r
symbols <- c("diamond", "circle")
```





####################################
Generating dds for M/F WB comparison
####################################
```{r pressure6, echo=FALSE}
dds <- DESeqDataSetFromMatrix( 
  countData = final_counts, 
  colData = All.coldata,
  design = ~ Batch + #explicitly incorporate sequencing batch into design matrix
  Tissue.Type
  ) 

design(dds)
colData(dds)

#Prefilter
smallestGroupSize <- 5
#1 -> smallest group 1 M or F
#2 -> smallest group size for female epidermal tissue.type samples
#10 -> smallest group for hemato vs not hemato
nc <- counts(dds, 
             normalized=F)
keep <- rowSums(nc >= 10) >= smallestGroupSize
dds <- dds[keep,]


#relevel dds to get proper coefficients of M.WB vs F.WB
#need to relevel dds <- https://www.biostars.org/p/448959/#484944 AND https://www.biostars.org/p/9464732/
#for geneList of GSEA to work
dds$Tissue.Type = relevel(dds$Tissue.Type, ref = "F.WB")
dds <- estimateSizeFactors(dds)
dds <- estimateDispersions(dds)
dds <- nbinomWaldTest(dds, maxit=500) #<- ALWAYS run after running bottom code
resultsNames(dds)

#DESeq dataset
dds <- DESeq(dds)
nc <- counts(dds, 
             normalized=T)
keep <- rowSums(nc >= 10) >= smallestGroupSize
dds <- dds[keep,]
dds <- DESeq(dds)

#Transformed matrices 
#this is in PCA section but needs to be rerun for "tissue.type" comparison
vsd <- vst(dds, 
           blind = T
           )
mat.vsd <- assay(vsd)
mm.vsd <- model.matrix(~Tissue.Type, colData(vsd)) #Sex #Tissue.Type
mat.vsd <- limma::removeBatchEffect(mat.vsd, 
                                    batch=vsd$Batch,  
                                    design=mm.vsd)
assay(vsd) <- mat.vsd

#examine post-batch correct via PCA
plotPCA(vsd,intgroup=c("Batch"))
plotPCA(vsd,intgroup=c("Organ.Type"))
plotPCA(vsd,intgroup=c("Tissue.Type"))
```





###################
Variance Testing###
###################
- 3.1.2026
  - VB says to checkout some variance in data with variancePartitioner
    - https://bioconductor.posit.co/packages/3.23/bioc/vignettes/variancePartition/inst/doc/variancePartition.html#input-data

- in the above block, we processed and normalized the gene-level counts, removed batch effect, and re-leveled dds before running variancePartition analysis.
```{r}
#Detecting problems caused by col-linearity of variables
form <- ~ Batch + Sex + Organ.Type #+ Tissue.Type
#form <- ~ Batch + Tissue.Type
#3.2.2026 - had an issue with colin. score = 1
#dropped hemato

# fit model
res <- fitVarPartModel(mat.vsd[1:500, ], form, All.coldata)

# evaluate the collinearity score on the first model fit
# this reports the correlation matrix between coefficient estimates
# for fixed effects
# the collinearity score is the maximum absolute correlation value
# If the collinearity score > .99 then the variance partition
# estimates may be problematic
# In that case, a least one variable should be omitted
colinearityScore(res[[1]])

#run vp
vp <- fitExtractVarPartModel(mat.vsd, form, All.coldata)

# sort variables (i.e. columns) by median fraction
#       of variance explained
vp <- sortCols(vp)


#plotting

# Bar plot of variance fractions for the first 10 genes
plotPercentBars(vp[1:10, ])

# violin plot of contribution of each variable to total variance
plotVarPart(vp)

# Compute Canonical Correlation Analysis (CCA)
# between all pairs of variables
# returns absolute correlation value
C <- canCorPairs(form, All.coldata)
# Plot correlation matrix
# between all pairs of variables
plotCorrMatrix(C)

```





##############
Results Tables
##############

M/W WB analysis (at the tissue type level)
```{r}
system.time({res <- results(dds,
                           independentFiltering = TRUE,
                           alpha = 0.05, #run .01 for M/F
                           contrast=c("Tissue.Type", "M.WB", "F.WB"),
                           lfcThreshold = 0 
)
})
summary(res)
head(res)

res.df <- as.data.frame(res)

#Shrinking LFC values - see "Making GeneList" in GSEA analysis below for clusterProfiler
#use lfcShrink <- https://support.bioconductor.org/p/9154747/
#need to relevel dds <- https://www.biostars.org/p/448959/#484944 AND https://www.biostars.org/p/9464732/
#releveled at beginning of pipeline to ensure I have F.WB as reference 
#resultsNames(dds) <- Look for the coefficient "Tissue.Type_M.WB_vs_F.WB"
resLFC <- lfcShrink(dds, 
                    coef = "Tissue.Type_M.WB_vs_F.WB",
                    res = res, #can use res here once dds is releveled originally 
                    type="apeglm")
summary(resLFC)
#subsetting DEGs with p.adj < .05
resLFC.Padj.df <- as.data.frame(resLFC) %>% #make df
  na.omit() %>% #remove NA
  subset(padj < .05) %>% #subset DEGs
  dplyr::arrange(desc(log2FoldChange)) #LFC values in descending order

#12/2/24 - making ordered l2fc matrix and padj df for heatmap annotation below
#getting and sorting shrunken LFC values - read this into the LFC annotation for heatmap below
resLFC.LFC <- resLFC.Padj.df[c("log2FoldChange")]
colnames(resLFC.LFC)[1] ="L2FC"
resLFC.LFC.mtx <- as.matrix(resLFC.LFC)

resLFC.Padj <- resLFC.Padj.df[c("padj")]
colnames(resLFC.LFC)[1] ="L2FC"



#Shrunken LEFC results table for ALL genes
res.sorted <- resLFC
#And for DEGs
res.sorted <- resLFC.Padj.df
res.df <- as.data.frame(res.sorted)

#writing to file
write.csv(res.df, "MWB_FWB_DESeq2_Res.csv")
```



Comparing Hemato (combined WBs) to other organ systems (at organ type level)
- remember, need to rerun dds object creation at organ type level and then relevel the factors on hemato (ie hemato should be reference - it's on epidermal cause if not specified, it's alphabetical)
```{r}

#making DESeq object
dds <- DESeqDataSetFromMatrix( 
  countData = final_counts, 
  colData = All.coldata, 
  design = ~ Batch + 
    Sex + 
    Organ.Type 
) 

design(dds)
colData(dds)

#Prefilter
smallestGroupSize <- 7 #7 -> smallest group size for symbiotic organ.type
keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
dds <- dds[keep,]

#relevel on hemato
dds$Organ.Type = relevel(dds$Organ.Type, ref = "Hematopoietic")
dds <- estimateSizeFactors(dds)
dds <- estimateDispersions(dds)
dds <- nbinomWaldTest(dds, maxit=500) #<- ALWAYS run after running bottom code
resultsNames(dds)

#DESeq algorithm
dds <- DESeq(dds)
nc <- counts(dds, 
             normalized=T)
keep <- rowSums(nc >= 10) >= smallestGroupSize
dds <- dds[keep,]
dds <- DESeq(dds)

#transformation
vsd <- vst(dds, 
           blind = T)

#batch correction
mat.vsd <- assay(vsd)
mm.vsd <- model.matrix(~Tissue.Type, colData(vsd)) #Sex #Tissue.Type
mat.vsd <- limma::removeBatchEffect(mat.vsd, 
                                    batch=vsd$Batch,  
                                    design=mm.vsd)
assay(vsd) <- mat.vsd

#examine post-batch correct via PCA
plotPCA(vsd,intgroup=c("Batch"))
plotPCA(vsd,intgroup=c("Organ.Type"))
```

And now hemato results tables and DEG extraction 
```{r}
system.time({res <- results(dds,
                           independentFiltering = TRUE,
                           alpha = 0.01, #run .01 for M/F
                           contrast=c("Organ.Type", "Epidermal", "Hematopoietic"),
                           lfcThreshold = 2
)
})
summary(res)
head(res)

res.df <- as.data.frame(res)

#shrink LFC
resLFC <- lfcShrink(dds,
                    coef = "Organ.Type_Epidermal_vs_Hematopoietic",
                    res = res, #can use res here once dds is releveled originally 
                    type="apeglm")
#Organ.Type_Symbiotic_vs_Hematopoietic
#Organ.Type_Gonadal_vs_Hematopoietic
#Organ.Type_Neural_vs_Hematopoietic
#Organ.Type_Epidermal_vs_Hematopoietic
summary(resLFC)


#subsetting DEGs
resLFC.Padj.df <- as.data.frame(resLFC) %>% #make df
  na.omit() %>% #remove NA
  subset(padj < .01) %>% #subset DEGs
  dplyr::arrange(desc(log2FoldChange)) #LFC values in descending order

#getting DEG gene lists for each organ type
Epidermal_DEGs <- rownames(resLFC.Padj.df %>%
                         subset(log2FoldChange > 0))
Hemato_v_Epidermal <- rownames(resLFC.Padj.df %>%
                         subset(log2FoldChange < 0))
#Symbioitc_DEGs
#Gonadal_DEGs
#Neural_DEGs
#Epidermal_DEGs

#writing organ type degs and res to file
write.table(Epidermal_DEGs, file = "Epidermal_v_Hemato_DEGs.txt", 
            sep="\t",
            col.names = F, 
            row.names = F,
            quote = FALSE,
)
#Symbio_v_Hemato_DEGs.txt
#Gonadal_v_Hemato_DEGs.txt
#Neural_v_Hemato_DEGs.txt

write.csv(resLFC.Padj.df, "Epidermal_v_Hemato.csv")
#Symbio_v_Hemato.csv
#Gonadal_v_Hemato.csv
#Neural_v_Hemato.csv

#combining wb degs and then making nunique
Comb_Hemato <- c(
  Hemato_v_Symbioitc,
  Hemato_v_Gonadal,
  Hemato_v_Neural,
  Hemato_v_Epidermal
) 
Comb_unique_hemato <- unique(Comb_Hemato)

#write final unique hemato gene list
write.table(Comb_unique_hemato, file = "Unique_hemato_degs.txt",
            sep="\t",
            col.names = F, 
            row.names = F,
            quote = FALSE)

```





#######
DEGs
#######
``` {r}

#extracting DEGs
DEGs.res <- subset(res.sorted, 
                   padj < 0.05, 
                   log2FoldChange > 0 | log2FoldChange < 0
)
DEGs.res.df <- as.data.frame(DEGs.res)
DEGs <- rownames(subset(DEGs.res))
DEGs.df <- as.data.frame(DEGs)
#write.table(DEGs, file = "M.F_WB_DEGs.txt",
            #row.names = F, col.names = F,
            #quote = F)

#Up and down regulated genes
Up.DEGs.res.df <- subset(res.sorted,
                         padj < 0.05 & 
                         log2FoldChange > 0) %>%
  as.data.frame()
Up.DEGs <- rownames(Up.DEGs.res.df)
Down.DEGs.res.df <- subset(res.sorted,
                           padj < 0.05 & 
                           log2FoldChange < 0) %>%
  as.data.frame()
Down.DEGs <- rownames(Down.DEGs.res.df)
#To extract logFold2changes and And p values
res.LFC.Padj <- subset(res.df, 
                            select=c("log2FoldChange", "padj"))
Up.DEGs.res.LFC.Padj <- subset(Up.DEGs.res.df, 
                               select=c("log2FoldChange", "padj"))
Down.DEGs.res.LFC.Padj <- subset(Down.DEGs.res.df, 
                                 select=c("log2FoldChange", "padj"))


#Transformed matrices of DEGs
vsd.DEGs <- vsd[DEGs,] %>% assay
vsd.Up.dge <- vsd[Up.DEGs,] %>% assay
vsd.Down.dge <- vsd[Down.DEGs,] %>% assay

```


- making non-ribo, non-uncharacterized DEG list 
```{r}
#filtering out ribosomal genes
DE.GOIs <- c(Up.DEGs, Down.DEGs)
df <- as.data.frame(DE.GOIs) #converting matrix to  
names(df)[names(df) == 'DE.GOIs'] <- 'name' ##renaming column
df_no_ribo = as.data.frame(df[!grepl("ribosom", df$name),]) #deleting rows with 'ribosom'
df_no_ribo_no_unchar = as.data.frame(df_no_ribo[!grepl("unchar", df_no_ribo$`df[!grepl("ribosom", df$name), ]`),]) #deleting rows with 'unchar'
Named.DE.GOIs <- df_no_ribo_no_unchar$`df_no_ribo[!grepl("unchar", df_no_ribo$\`df[!grepl("ribosom", df$name), ]\`), ]`#making new Named.DEGs variable from these gene names
```


Finding top variable WB genes (Fig. 3)
```{r}
#and Top 100 variable WB genes
select <- order(rowMeans(counts(dds[,1:10],normalized=TRUE)), #dds[,1:10] is for top 100 WB genes 
                              decreasing=TRUE)[1:100]
vsd.select <- vsd[select,] %>% assay()
#write.table(rownames(vsd.select),
#file = "/Users/johnbriseno/Desktop/top_100 _WB_genes",
#quote = F,
#row.names = F, col.names = F)

#filtering out ribosomal genes
df <- as.data.frame(vsd.select) #converting matrix to  
df$name <- rownames(df) #making new 'name' column w the rownames to drop
df_no_ribo = df[!grepl("ribosom", df$name),] #deleting rows with 'ribosom'
df_no_ribo_no_unchar = df_no_ribo[!grepl("unchar", df_no_ribo$name),] #deleting rows with 'ribosom'
drops = 'name'
df_no_ribo_no_unchar = df_no_ribo_no_unchar[ , !names(df_no_ribo_no_unchar) %in% drops] #dropping 'name' column
#looks like theres 81 non ribo genes, but gonna just show top 50 genes (rows)
top_df_no_ribo_no_char = df_no_ribo_no_unchar[1:50,]
mtx_no_ribo_no_unchar = data.matrix(top_df_no_ribo_no_char) #reconverting to mtx for heatmap



#OL for VB from 2025
OL.select <- order(rowMeans(counts(dds[,17:22],normalized=TRUE)), #dds[,1:10] is for top 100 WB genes 
                              decreasing=TRUE)[1:100]
vsd.OL.select <- vsd[OL.select] %>% assay()




#To JUST get WBs in Heatmaps section like model fig 7
WB.vsd.DEGs <- vsd.DEGs[,1:10]

#have to reorder 
#https://stackoverflow.com/questions/1568511/how-do-i-sort-one-vector-based-on-values-of-another
#Named.DE.GOIs <- Named.DE.GOIs[order(match(Named.DE.GOIs, rownames(VSD.padj.DEGs.df)))]
#DEGs <- DEGs[order(match(DEGs, rownames(VSD.padj.DEGs.df)))]
#DE.GOIs <- DE.GOIs[order(match(DE.GOIs, rownames(VSD.padj.DEGs.df)))]

```





####################
Genes of Interest###
####################
- This ones gonna take some time 
- Searching NCBI of gene names and seeing what sticks
- use 'final_counts' to determine counts since assay(vsd) has this normalized (i.e. low values)
- Sets up gene names for Hemato/Immuno Fig. 4 and WB progression model of Fig. 7


The Role of Hemocytes (McAnultuy & Nyholm, 2017) homologs:
```{r}

CDs <- 
  c(
    "LOC145288272_CD109_antigen-like"
    #"LOC145288265_CD109_antigen-like",
    #"LOC145293977_CD109_antigen-like",
    #"LOC145288327_CD109_antigen-like",
    #"LOC145241389_CD109_antigen-like"
  )

SUSHI.Repeat <- c(
  "LOC145249753_sushi,_von_Willebrand_factor_type_A,_EGF_and_pentraxin_domain-containing_protein_1-like"#,
  #"LOC145291698_sushi,_von_Willebrand_factor_type_A,_EGF_and_pentraxin_domain-containing_protein_1-like"
)

PGRPs <- c(
  "LOC145245935_peptidoglycan_recognition_protein_1-like", #PGRP1
  "LOC145245936_peptidoglycan_recognition_protein_1-like", #PGRP2
  "LOC145245938_peptidoglycan_recognition_protein_1-like", #PGRP3
  "LOC145245940_peptidoglycan_recognition_protein_1-like", #PGRP4
  "LOC145291335_EsPGRP5" #PGRP5
)

NFkB.Molecules <- 
  c(
    
    #Tolls
    #"LOC145291538_toll-like_receptor_2_type-2",
    #"LOC145243454_toll-like_receptor_2",
    #"LOC145265434_toll-like_receptor_2",
    #"LOC145252737_toll-like_receptor_4",
    "LOC145254925_toll-like_receptor_4",
    #"LOC145258232_toll-like_receptor_Tollo",
    #"LOC145291122_toll-like_receptor_Tollo",
    #"LOC145291547_toll-like_receptor_Tollo",
    #"LOC145285835_toll-interacting_protein-like",
    
    #IRAKs
    "LOC145243982_interleukin-1_receptor-associated_kinase_4-like",
    #"LOC145294871_interleukin-1_receptor-associated_kinase_4-like",
    #"LOC145247953_interleukin-1_receptor-associated_kinase_1-binding_protein_1_homolog",
    "LOC145267169_interleukin_enhancer-binding_factor_2_homolog",
    
    #TRAFs
    "LOC145284867_tumor_necrosis_factor_alpha-induced_protein_8-like",
    #"LOC145248052_lipopolysaccharide-induced_tumor_necrosis_factor-alpha_factor_homolog",
    #"LOC145265618_lipopolysaccharide-induced_tumor_necrosis_factor-alpha_factor_homolog",
    
    #NfKB/Rel
    "LOC145293216_nuclear_factor_NF-kappa-B_p105_subunit-like",
    #"LOC145278029_nuclear_factor_NF-kappa-B_p105_subunit-like",
    #"LOC145270024_NF-kappa-B_inhibitor-interacting_Ras-like_protein_1",
    "LOC145260317_NF-kappa-B_inhibitor_alpha-like",
    #"LOC145252681_inhibitor_of_nuclear_factor_kappa-B_kinase_subunit_alpha-like",
    
    #Myd88
    "LOC145259775_myeloid_differentiation_primary_response_protein_MyD88-like"
  )

Histones <- c(
  "LOC145266374_histone_H3.3A",
  "LOC145249034_histone_H4_transcription_factor-like",
  "LOC145249510_histone_H5-like"
)

Cathepsin <- c(
  "LOC145279650_cathepsin_B-like",
  "LOC145245986_cathepsin_L-like_peptidase",
  "LOC145265457_cathepsin_Z-like"
)

Galectin <- c(
  "LOC145279619_galectin-3-binding_protein_A-like",
  "LOC145260397_galectin-4-like",
  "LOC145293841_galectin-4-like"
)

Cephalotoxin <- c(
  #"LOC145248062_SE-cephalotoxin-like"
)

Chitotriosidase <- c(
  #"LOC145279759_chitotriosidase-1-like",
  #"LOC145292182_chitotriosidase-1-like",
  "LOC145293110_chitotriosidase-1-like"
)

Immuno.Seqs <- c(PGRPs,
                 Galectin,
                 NFkB.Molecules,
                 #Cephalotoxin,
                 Chitotriosidase,
                 Histones,
                 Cathepsin,
                 CDs,
                 SUSHI.Repeat
)

Immuno_annotation_row = data.frame(
  Hemocyte.Genes = factor(rep(c("PGRPs","Galectin","NFkB.Molecules",#"Cephalotoxin"
                                "Chitotriosidase","Histones","Cathepsin","CDs","SUSHI.Repeat"
                                ),
                              c(
                                #for just WB showing in Fig. 4B
                                5, #PGRPs
                                3, #Galectin
                                7, #NFkB
                                1, #Chitotriosidase
                                3, #histones
                                3, #cathepsin
                                1, #CD109
                                1 #SUSHI
                                
                                #for full list in Fig. S5B
                                #5, #PGRPs,
                                #3, #Galectin
                                #22, #NFkB.Molecules
                                #0, #Cephalotoxin
                                #3, #Chitotriosidase
                                #3, #histones
                                #3, #cathepsin
                                #1, #CDs
                                #2 #SUSHI Repeat
                              ))))
rownames(Immuno_annotation_row) = Immuno.Seqs

#Am deciding sequence order before hand
#refactoring to get ID
Immuno.Seqs.Groups <- factor(Immuno_annotation_row$Hemocyte.Genes,
levels = c("PGRPs","Galectin","NFkB.Molecules",
           "Chitotriosidase","Histones","Cathepsin","CDs","SUSHI.Repeat"
))

#Adding color 
Immuno_row_colors = list(
  Hemocyte.Genes = c(
    
    PGRPs = "#009E73",
    Galectin = "#009E73",
    
    NFkB.Molecules = "red3",
    
    Chitotriosidase = "orange",
    Histones = "orange",
    Cathepsin = "orange",
    
    CDs = "skyblue3",
    SUSHI.Repeat = "skyblue3"
    )
)


```



Hematopoietic homologs:
```{r}
Hemato_Stem <- c(
  "LOC145267028_HHEX*",
  "LOC145291535_RUNX1*",
  
  "LOC145259310_ETS-related_transcription_factor_Elf-2-like",
  
  #"LOC145258217_GATA_zinc_finger_domain-containing_protein_1-like",
  #"LOC145280112_GATA-type_zinc_finger_protein_1-like",
  "LOC145296204_transcription_factor_GATA-3-like"
)
Integrins <- c(
  #"LOC145263070_integrin_alpha-9-like",
  "LOC145291142_integrin_alpha-6-like",
  "LOC145291306_integrin_alpha-9-like",
  "LOC145291435_integrin_alpha-9-like",
  "LOC145292146_integrin_alpha-9-like"#,
  #"LOC145292160_integrin_alpha-4-like",
  #"LOC145292206_integrin_alpha-8-like"
)
CDs <- c(
  "LOC145269126_CD2_antigen_cytoplasmic_tail-binding_protein_2_homolog",
  
  #"LOC145292897_CD9_antigen-like",
  #"LOC145293212_CD9_antigen-like",
  "LOC145293676_CD9_antigen-like",
  #"LOC145293677_CD9_antigen-like",
  #"LOC145293678_CD9_antigen-like",
  
  #"LOC145241564_CD63_antigen-like",
  #"LOC145247155_CD63_antigen-like",
  #"LOC145247156_CD63_antigen-like",
  #"LOC145247158_CD63_antigen-like",
  #"LOC145247159_CD63_antigen-like",
  #"LOC145247248_CD63_antigen-like",
  #"LOC145247956_CD63_antigen-like",
  #"LOC145252853_CD63_antigen-like",
  "LOC145292904_CD63_antigen-like",
  #"LOC145280157_early_activation_antigen_CD69-like",
  
  #"LOC145241389_CD109_antigen-like",
  #"LOC145288265_CD109_antigen-like",
  "LOC145288272_CD109_antigen-like",
  #"LOC145288327_CD109_antigen-like",
  #"LOC145293977_CD109_antigen-like",
  
  "LOC145252526_CD151_antigen-like",
  "LOC145259315_CD151_antigen-like"
)

Lymphoid <- c(
  #"LOC145239861_mucosa-associated_lymphoid_tissue_lymphoma_translocation_protein_1-like", #MALT1
  #"LOC145256258_lymphotoxin_beta_receptor_inhibitor-like", #LTβR
  "LOC145284740_lymphocyte-specific_helicase-like"#, #HELLS
  #"LOC145277010_cryptochrome-1-like", #CRY1
  #"LOC145293881_cryptochrome-1-like"
)

B_cell <- c(
  #"LOC145239864_B-cell_lymphoma/leukemia_10-like", #BCL10
  #"LOC145288011_B-cell_lymphoma_3_protein-like", #BCL-3
  "LOC145256486_B-cell_receptor-associated_protein_31-like"#, #BCAP31
  #"LOC145268288_B-cell_receptor-associated_protein_31-like", #BCAP31
  #"LOC145268799_pre-B-cell_leukemia_transcription_factor_1-like" #PBX1
)

T_cell <- c(
  #"LOC145268813_T-cell_activation_inhibitor,_mitochondrial-like", #TCAIM or TOAG1
  #"LOC145287939_malignant_T-cell-amplified_sequence_1-like", #MCTS1 or MCT-1
  "LOC145296256_T-cell_immunomodulatory_protein-like" #TIP or ITFG1
)

Myeloid <- c(
  #"LOC145259775_myeloid_differentiation_primary_response_protein_MyD88-like", #MYD88
  "LOC145276911_induced_myeloid_leukemia_cell_differentiation_protein_Mcl-1_homolog"#, #MCL1
  #"LOC145297776_myeloid-derived_growth_factor-like", #MYDGF
  #"LOC145297803_myeloid_leukemia_factor_1-like", #MLF1
  #"LOC145303596_myeloid_leukemia_factor_2-like" #MLF2
  #X#"LOC145303598_myeloid_leukemia_factor_2-like" #MLF2
)

Platelet <- c(
  "LOC145277628_platelet-activating_factor_acetylhydrolase_IB_subunit_alpha2-like"#, #PAFAH1B2
  #"LOC145279758_platelet-activating_factor_acetylhydrolase-like" #PAF-AH
)

Macrophage <- c(
  #"LOC145254682_monocyte_to_macrophage_differentiation_factor_2-like", #MMD2
  #"LOC145269864_macrophage_metalloelastase-like", #MMP
  "LOC145277553_macrophage_migration_inhibitory_factor-like", #MIF
  #"LOC145277554_macrophage_migration_inhibitory_factor-like", #MIF
  "LOC145284844_natural_resistance-associated_macrophage_protein_2-like"#, #NRAMP2 or DMT1 or SLC11A2
  #"LOC145291322_natural_resistance-associated_macrophage_protein_2-like" ##NRAMP2 or DMT1 or SLC11A2
)

Leukocyte <- c(
  #"LOC145243966_leukocyte_tyrosine_kinase_receptor-like",
  #"LOC145252475_leukocyte_elastase_inhibitor-like",
  #"LOC145252476_leukocyte_elastase_inhibitor_A_pseudogene",
  #"LOC145252583_leukocyte_receptor_cluster_member_9-like",
  #"LOC145261077_cysteinyl_leukotriene_receptor_1-like",
  #"LOC145269125_leukocyte_receptor_cluster_member_1_homolog",
  #X#"LOC145274194_leukotoxin_export_ATP-binding_protein_LtxB-like",
  "LOC145284903_leukocyte_receptor_cluster_member_8_homolog",
  "LOC145288069_leukocyte_elastase_inhibitor-like"#,
  #"LOC145297606_leukotriene_A-4_hydrolase-like"
)

Melanocyte <- c(
  "LOC145248024_leucine-rich_melanocyte_differentiation-associated_protein-like",
  "LOC145288655_leucine-rich_melanocyte_differentiation-associated_protein-like"
)

Erythrocyte <- c(
  "LOC145252556_5-aminolevulinate_synthase,_erythroid-specific,_mitochondrial-like"
)

NKcells <- c(
  "LOC145266719_killer_cell_lectin-like_receptor_subfamily_F_member_1"
)

Endothelial <- c(
  #"LOC145241505_endothelial_differentiation-related_factor_1_homolog",
  #"LOC145242090_endothelin-converting_enzyme_homolog",
  #"LOC145254767_BMP-binding_endothelial_regulator_protein-like",
  "LOC145258150_vascular_endothelial_growth_factor_receptor_1-like",
  #"LOC145261200_endothelin-converting_enzyme_homolog",
  "LOC145270282_endothelin-converting_enzyme_1-like"
)

Fibroblast <- c(
  #"LOC145242616_fibroblast_growth_factor_receptor-like_1",
  #"LOC145247897_fibroblast_growth_factor_receptor_3-like",
  #"LOC145248700_fibroblast_growth_factor_receptor_2-like",
  "LOC145263773_fibroblast_growth_factor_receptor_1-like"#,
  #"LOC145266927_fibroblast_growth_factor_18-like",
  #"LOC145277826_fibroblast_growth_factor_9-like",
  #"LOC145283546_fibroblast_growth_factor_receptor_3-like",
  #"LOC145292938_acidic_fibroblast_growth_factor_intracellular-binding_protein_B-like"
)

Cytokine <- c(
  #"LOC145268228_cytokine-like_nuclear_factor_N-PAC",
  #"LOC145285903_suppressor_of_cytokine_signaling_2-like",
  #"LOC145291943_suppressor_of_cytokine_signaling_2-like",
  "LOC145294080_cytokine_receptor-like_factor_3"
)

Hemato.Seqs <- c(
  Hemato_Stem, CDs, Integrins, 
    
  Endothelial, Fibroblast, Cytokine,
  
  Lymphoid, B_cell, T_cell, NKcells,
  
  Myeloid, Platelet, Macrophage, Leukocyte, Melanocyte, Erythrocyte

)

#check what genes are not in transformed dataset
missing_rows <- Hemato.Seqs[ !Hemato.Seqs %in% rownames(assay(vsd)) ]
print(missing_rows)

Hemato_annotation_row = data.frame(
                Hemato.Gene.Class = factor(rep(c(
                  "Hemato_Stem", "CDs", "Integrins", 
                  "Endothelial", "Fibroblast", "Cytokine",
                  "Lymphoid", "B_cell", "T_cell","NKcells",
                  "Myeloid", "Platelet", "Macrophage", "Leukocyte", "Melanocyte", "Erythrocyte"
                  ),
                  c(
                    #Partial list - Fig. 4A
                    4, #hemato
                    6, #CDs
                    4, #integrins
                    2, #endo
                    1, #fibro
                    1, #cyto
                    1, #lymphoid
                    1, #Bcell
                    1, #TCell
                    1, #NK
                    1, #Myeloid
                    1, #Platelet
                    2, #macro
                    2, #leuko
                    2, #melano
                    1 #erythrocyte
                    
                    #full list - Fig. S5A
                    #6, #hemato
                    #16, #CDs
                    #7, #Integrins
                    
                    #6, #Endothelial
                    #8, #Fibroblast
                    #4, #cytokine
                    
                    #5, #Lymphoid
                    #5, #B_cell
                    #3, #T-cell
                    #1, #NKcells
                    
                    #4, #Myeloid
                    #2, #Platelet
                    #6, #macrophage
                    #9, #Leukocyte
                    #2, #melanocyte
                    #1 #Erythrocyte
                    ))))
rownames(Hemato_annotation_row) = Hemato.Seqs

#Am deciding sequence order before hand
#refactoring to get ID

Hemato.Seqs.Groups <- factor(Hemato_annotation_row$Hemato.Gene.Class,
                             levels = c("Hemato_Stem", "CDs", "Integrins",
                                        "Endothelial", "Fibroblast", "Cytokine",
                                        "Lymphoid", "B_cell", "T_cell", "NKcells",
                                        "Myeloid", "Platelet", "Macrophage", "Leukocyte", "Melanocyte", "Erythrocyte" 
                             ))

#Adding color 
Hemato_row_colors = list(
  Hemato.Gene.Class = c(
                               Hemato_Stem = "red1",
                               CDs = "red4",
                               Integrins = "red3",
                               
                               Endothelial = "indianred3",
                               Fibroblast = "indianred4",
                               Cytokine = "cyan2",

                               
                               Lymphoid = "khaki2",
                               B_cell = "gold2",
                               T_cell = "goldenrod1",
                               NKcells = "yellow4",
                               
                               Myeloid = "dodgerblue",
                               Platelet = "lightblue3",
                               Macrophage = "royalblue",
                               Leukocyte = "forestgreen",
                               Melanocyte = "blue1",
                               Erythrocyte = "blue4"
))

```



WB GOIs for model
```{r}
sn.genes <- c(
  
  #TFs
  "LOC145267028_HHEX*",
  "LOC145291535_RUNX1*",
  "LOC145296204_transcription_factor_GATA-3-like",
  "LOC145249753_sushi,_von_Willebrand_factor_type_A,_EGF_and_pentraxin_domain-containing_protein_1-like",
  "LOC145259310_ETS-related_transcription_factor_Elf-2-like",
  
  #HNRNPs
  "LOC145255650_heterogeneous_nuclear_ribonucleoproteins_A2/B1-like",
  "LOC145255648_heterogeneous_nuclear_ribonucleoproteins_A1_homolog",
  "LOC145258173_heterogeneous_nuclear_ribonucleoprotein_A1-like_3",
  "LOC145260563_heterogeneous_nuclear_ribonucleoprotein_U-like_protein_1",
  "LOC145276714_heterogeneous_nuclear_ribonucleoprotein_K-like",
  
  #Actin-mediated cytoplasmic bridge
  "LOC145259195_actin,_cytoplasmic_1-like",
  "LOC145241409_actin-related_protein_5-like",
  "LOC145243680_coactosin-like_protein",
  "LOC145247564_actin-related_protein_2/3_complex_subunit_5-B-like",
  "LOC145245752_kinesin-like_protein_KIF11-A",
  
  #Ras gtpase
  "LOC145252733_rho_GDP-dissociation_inhibitor_2-like",
  "LOC145260529_ras-like_GTP-binding_protein_RHO",
  "LOC145270024_NF-kappa-B_inhibitor-interacting_Ras-like_protein_1",
  "LOC145239760_GTPase_KRas-like",
  "LOC145263723_ras-related_protein_M-Ras-like",
  
  #PRRs
  "LOC145291335_EsPGRP5",
  "LOC145266610_scavenger_receptor_cysteine-rich_domain_superfamily_protein-like",
  "LOC145296256_T-cell_immunomodulatory_protein-like",
  "LOC145254925_toll-like_receptor_4",
  "LOC145279619_galectin-3-binding_protein_A-like",
  
  #Iron regulation
  "LOC145283795_hephaestin-like_protein",
  "LOC145246871_ferritin-1_heavy_chain-like",
  "LOC145242702_soma_ferritin-like",
  "LOC145254756_iron-sulfur_cluster_transfer_protein_NUBPL-like",
  "LOC145241961_succinate_dehydrogenase_[ubiquinone]_iron-sulfur_subunit,_mitochondrial-like"
)



```



########
HeatMaps
########
Setup happens here for columns/rows annotations 
As well as LFC/p.Adj values

```{r}
#row annotations for specific groups of GOIs are in the above GOIs section

#column annotation
annotation_col = data.frame((colData(dds)[("Organ.Type")]),
                            (colData(dds)[("Sex")]),
                            (colData(dds)[("Tissue.Type")]),
                            (colData(dds)[("Hemato")])
)
```





#################
HeatMaps for DEGs
#################
- Code for: 
    - Fig. 5 - Sex-Specific WB DEGs
    - Figs. S# - Full sex-specific WB DEG list across tissues

- Adding in LFC/p.Adj values (row annotations)
```{r}
#12/2/24 - rewrote adding lfc/padj annots to just pull out ordered l2fc/padj and make it a matrix/df after i made results shrink table

######################
#adding lfc somehow###
######################
resLFC.LFC.mtx
#pulling out DEG subset
VSD.LFC.DEGs.mtx <- subset(resLFC.LFC.mtx, 
                                 rownames(resLFC.LFC.mtx) %in% DEGs)
VSD.LFC.Named.DEGs.mtx <- subset(resLFC.LFC.mtx, 
                                 rownames(resLFC.LFC.mtx) %in% Named.DE.GOIs)
#pull out values greater than LFC = |.5|
# Keep rows where EVERY element is > 0.5 OR < -0.5
filtered_rows <- VSD.LFC.Named.DEGs.mtx[rowSums(abs(VSD.LFC.Named.DEGs.mtx) > 0.5) == ncol(VSD.LFC.Named.DEGs.mtx), , drop = FALSE]
filt_VSD.LFC.Named.DEGs.mtx = filtered_rows


######################################
#adding padj info to heatmap somehow#
######################################
resLFC.Padj
colnames(resLFC.Padj)[1] ="P.Adj"
#pulling out named DEG subset
VSD.padj.DEGs.df <- subset(resLFC.Padj,
                                 rownames(resLFC.Padj) %in% DEGs)
VSD.padj.Named.DEGs.df <- subset(resLFC.Padj,
                                 rownames(resLFC.Padj) %in% Named.DE.GOIs)
#extracting rows that match the subsetted VSD.LFC.Named.DEGs.mtx
df = VSD.padj.Named.DEGs.df
filt_degs <- rownames(filt_VSD.LFC.Named.DEGs.mtx)
filt_df <- VSD.padj.Named.DEGs.df[rownames(VSD.padj.Named.DEGs.df) %in% filt_degs, , drop = FALSE]
filt_VSD.padj.Named.DEGs.df = filt_df

#first extract filt Named DEGs 
# Filter to KEEP items present in the filter vector
filt_Named_DEGs <- Named.DE.GOIs[Named.DE.GOIs %in% filt_degs]


#Need reorder gene subset in the order of the LFC values
#https://stackoverflow.com/questions/17031039/how-to-sort-a-character-vector-according-to-a-specific-order
#Named.DE.GOIs <- Named.DE.GOIs[order(match(Named.DE.GOIs, rownames(VSD.LFC.DEGs.mtx)))]
#Named.DE.GOIs <- Named.DE.GOIs[order(match(Named.DE.GOIs, rownames(VSD.LFC.Named.DEGs.mtx)))]
```


- now making heatmap with row annotations 
```{r}
#heating the map for All DEGs
Candidates.Heatmap <- ComplexHeatmap::pheatmap(assay(vsd)[DEGs,],
                                               #assay(vsd)[Named.DE.GOIs,],
                                               #assay(vsd)[filt_Named_DEGs,1:10]
                                               #assay(vsd)[DEGs,],
                                               #vsd.DEGs
                                               #WB.vsd.DEGs
                                               #WB.vsd.DEGs[DE.GOIs,]
                                               
                                               cluster_rows=F, 
                                               cluster_cols=F,
                                               
                                               show_rownames=T,
                                               show_colnames = T,
                                               
                                               #Set up for Organ.Type
                                               column_split = annotation_col$Organ.Type,
                                               annotation_names_col = T,
                                               column_title = NULL,
                                               
                                               top_annotation = HeatmapAnnotation(Organ.Type = anno_block(gp = gpar(fill = c("#E18004", "#0072B2", "navajowhite", "yellow4", "#009E73")),
                                               labels = c("Epidermal (MNTL, SKIN, ARM)", 
                                               "Gonadal (OVRY, TST)", 
                                               "Hematopoietic (WB)", 
                                               "Neural (CB, OL)", 
                                               "Symbioitc (LO_CC)"),  
                                               labels_gp = gpar(col = "white", fontsize = 11))#,
                                               ),
                                               
                                               
                                               fontsize_row = 10,
                                               fontsize_col = 10,
                                               
                                               col = rev(brewer.pal(11, "RdBu")),
                                               border = NA,
                                               
                                               scale = "row",
                                               name = "Z-Score",
                                               #name = "VST Counts",
                                               #annotation_colors = ann_colors,
                                               heatmap_legend_param = list(legend_direction = "horizontal"
                                                                           #legend_width = unit(, "cm")
                                               ),
                                               #Horizontal Z scoring -> https://stackoverflow.com/questions/52794692/complexheatmap-cannot-create-horizontal-legend
                                               angle_col = "45",
                                               
                                               
                                               
                                               #add L2FC and p.adj
                                               #https://stackoverflow.com/questions/61050899/complexheatmap-increase-label-size-of-side-annotation
                                               #https://jokergoo.github.io/ComplexHeatmap-reference/book/heatmap-annotations.html#box-annotation
                                               #https://bioconductor.riken.jp/packages/3.8/bioc/vignettes/ComplexHeatmap/inst/doc/s4.heatmap_annotation.html
                                               right_annotation = rowAnnotation(#P.Adj Annot
                                                 df = VSD.padj.DEGs.df, 
                                                 #VSD.padj.Named.DEGs.df, 
                                                 #VSD.padj.DEGs.df
                                                 #filt_VSD.padj.Named.DEGs.df
                                                 col = list(P.Adj = colorRamp2(c(0, .01, .001), hcl_palette = "Viridis")
                                                 ),
                                                 #L2FC annot
                                                 Log2FC = anno_barplot(VSD.LFC.DEGs.mtx, 
                                                                       #VSD.LFC.DEGs.mtx
                                                                       #VSD.LFC.Named.DEGs.mtx
                                                                       #filt_VSD.LFC.Named.DEGs.mtx
                                                                       bar_width = .75,
                                                                       #width = unit(3, "cm"),
                                                                       gp = gpar(fill = ifelse(VSD.LFC.DEGs.mtx > 0,
                                                                                               #VSD.LFC.DEGs.mtx
                                                                                               #VSD.LFC.Named.DEGs.mtx
                                                                                               #filt_VSD.LFC.Named.DEGs.mtx
                                                                                               "deepskyblue", "deeppink2"),
                                                                                 fontsize = 10),
                                                                       axis_param = list(
                                                                         gp = gpar(fontsize = 8),
                                                                         labels_rot = 45)
                                                 ),
                                                 
                                                 #styling PAdj/L2FC legends
                                                 annotation_name_rot = c(0, 0),
                                                 annotation_width = unit(c(1.5, .7), "cm"),
                                                 gp = gpar(col = "black"),
                                                 show_annotation_name = c(P.Adj = FALSE),
                                                 annotation_legend_param = list(title_gp=gpar(fontsize=10,
                                                                                              fontface="bold"),
                                                                                labels_gp = gpar(fontsize = 10),
                                                                                direction = "horizontal")
                                               )
)
ComplexHeatmap::draw(Candidates.Heatmap, 
                     heatmap_legend_side="bottom", #left
                     annotation_legend_side = "bottom", #bottom
                     legend_grouping = "original",
                     padding = unit(c(2, 20, 2, 70), "mm")) ## see right heatmap in following)

```



#################
HeatMaps for GOIs
#################
```{r}
Candidates.Heatmap <- ComplexHeatmap::pheatmap(assay(vsd)[sn.genes,1:10],
                                               #Top WB expression figs (Fig. 3, Fig. S3)
                                               #mtx_no_ribo_no_unchar
                                               #vsd.select,
                                               
                                               
                                               #hemato/immune figs (Figs. 4,S5)
                                               #assay(vsd)[Immuno.Seqs,]
                                               #assay(vsd)[Immuno.Seqs,1:10]
                                               #assay(vsd)[Hemato.Seqs,]
                                               #assay(vsd)[Hemato.Seqs,1:10]
                                               
                                               #sn genes for model (Fig. 7)
                                               #assay(vsd)[sn.genes,]
                                               
                                               #old random genes of interest
                                               #vsd.OL.select
                                               #assay(vsd)[HCR_WB_GOIs,]
                                               
                                               

                                               cluster_rows=F, 
                                               #cluster_row_slices = T,
                                               cluster_cols=F,
                                               
                                               show_rownames=T,
                                               show_colnames = T,
                                               
                                               #Set up for Organ.Type
                                               #top_annotation = HeatmapAnnotation(Organ.Type = anno_block(gp = gpar(fill = c("#E18004", 
                                                                                                                             #"#0072B2", 
                                                                                                                             #"navajowhite", 
                                                                                                                             #"yellow4", 
                                                                                                                             #"#009E73")),
                                               #labels = c("Epidermal (MNTL, SKIN, ARM)", 
                                                #          "Gonadal (OVRY, TST)", 
                                                #          "Hematopoietic (WB)", 
                                                #          "Neural (CB, OL)", 
                                                #          "Symbioitc (LO_CC)"),  
                                               #labels_gp = gpar(col = "white", fontsize = 12))),
                                               
                                               
                                               #column annotation
                                               #column_split = annotation_col$Organ.Type,
                                               #annotation_names_col = F,
                                               #column_title = NULL,
                                               
                                               #row annotation
                                               #annotation_row = Hemato_annotation_row, 
                                               #Immuno_annotation_row #Hemato_annotation_row
                                               
                                               #row_split = Hemato.Seqs.Groups, 
                                               #Immuno.Seqs.Groups #Hemato.Seqs.Groups
                                               
                                               #annotation_names_row = F, #removes annotation label thank god #https://support.bioconductor.org/p/131791/
                                               #row_title_rot = 0,
                                               #row_title = NULL,
                                               #annotation_colors = Hemato_row_colors, #Immuno_row_colors, #Hemato_row_colors

                                               fontsize_row = 12,
                                               fontsize_col = 8,
                                               col = rev(brewer.pal(11, "PuOr")),#rev(),
                                               border = F,
                                               
                                               #Z scaling or transformed counts
                                               #scale = "row", #z score (observes per sample/#std deviations away from mean gene expression)
                                               #name = "Z-Score",
                                               name = "VST Counts",
                                               
                                               heatmap_legend_param = list(legend_direction = "horizontal"#,
                                                                           #legend_width = unit(, "cm")
                                               ),
                                               
                                               #Horizontal Z scoring -> https://stackoverflow.com/questions/52794692/complexheatmap-cannot-create-horizontal-legend
                                               angle_col = "45"
                                               )
                                          

ComplexHeatmap::draw(Candidates.Heatmap, 
     heatmap_legend_side="bottom", 
     annotation_legend_side = "bottom", 
     legend_grouping = "original",
     padding = unit(c(2, 2, 2, 120), "mm")) ## see right heatmap in following)
#save image size so it fits:
#like 1700 width/1800 height
#padding is good but need to make it more narrow
#and maybe for vst have names on left and swap dendrogram?

```






#############################################################
KEGG/GO Analysis with ClusterProfiler from EggNOG annotations
#############################################################

DESeq2 was used to compare: 
- M/F white bodies
- Pair-wise comparisons of upregulated WB genes across all tissues by comparing the hematopoietic white bodies to each organ type (epidermal, gonadal,neural, and symbiotic) independently then pooling the significant white body DEGs from these data. 
  - See the PCA section above for how the DDS object was generated for this

I have been unable to extract GO/KO terms EggNOG-annotated Preferred Names for gene IDs - 11/19/2024, I think I got it! Must extract GO/KOs when setting preferred gene names in beginning like so:

Reading in EnTAP
```{r}
EggNOG.KEGG.GO <- fread("clusterprofiler/entap_results.tsv",
                     select = c("Query_Sequence", 
                                "Database_EggNOG_KEGG_KO", 
                                "Database_EggNOG_GO_Biological",
                                "Database_EggNOG_GO_Cellular",
                                "Database_EggNOG_GO_Molecular"
                                ))
#have NaN's here from entap but probably shouldn't drop any
#some KOs are NaN whereas some GOs aren't and vice versa
```


Merging edited LOC IDs with entap
```{r}
#read in merged name df from top of this file
merged_df <- read.csv('NCBI_LOC_Descript_counts_matrix.csv')

#change col name to match entap col with LOC IDs
colnames(merged_df)[colnames(merged_df) == 'symbol'] <- 'Query_Sequence'

#merge
merge_LOC_entap <- merge(EggNOG.KEGG.GO, merged_df, by = "Query_Sequence", all.x=TRUE, all.y=TRUE)

#omit NAs
merge_LOC_entap <- na.omit(merge_LOC_entap)

#ensuring df structure
merge_LOC_entap <- as.data.frame(merge_LOC_entap)

#keeping KOs, GOs, new names
#https://stackoverflow.com/questions/4605206/drop-data-frame-columns-by-name
keeps <- c("Description",
           "Database_EggNOG_KEGG_KO",
           "Database_EggNOG_GO_Biological",
           "Database_EggNOG_GO_Cellular",
           "Database_EggNOG_GO_Molecular"
)
merge_LOC_entap <- merge_LOC_entap[ , keeps, drop = FALSE]
```


Dropping 'ribosomal' genes
```{r}
merge_LOC_entap <- merge_LOC_entap[!grepl("ribosom", merge_LOC_entap$Description),]
#drops about 4K genes
```


Making new columns with all GOS
- GOs are split by category (BP, CC, MF) which is ideal but they'll get split again not to worry!
- https://stackoverflow.com/questions/18115550/combine-two-or-more-columns-in-a-dataframe-into-a-new-column-with-a-new-name
```{r}
df = merge_LOC_entap
df$GOs <- paste(df$Database_EggNOG_GO_Biological,
                df$Database_EggNOG_GO_Cellular,
                df$Database_EggNOG_GO_Molecular,
                sep=","
                )
df$GOs <- gsub(",,", ",", df$GO) #changing double commas to single
```



Splitting by GO/KOs
```{r}

KO_df <- data.frame(df$Description,
                       df$Database_EggNOG_KEGG_KO)

GO_df <- data.frame(df$Description,
                       df$GOs)

#cleaning NaN finally
KO_df <- KO_df[!grepl("NaN", KO_df$df.Database_EggNOG_KEGG_KO),]

#have to be more through w the GOs NaN though
GO_df$df.GOs <- gsub("NaN,", "", GO_df$df.GOs) #removing NaN
GO_df$df.GOs <- gsub(",NaN", "", GO_df$df.GOs) #removing NaN
GO_df <- GO_df[!grepl("NaN", GO_df$df.GOs),]

#fixing col names
colnames(KO_df) <- c("ID", "KOs")
colnames(GO_df) <- c("ID", "GOID")
```



Saving final edited GO/KO lists
```{r}
write.table(GO_df, file = "LOC_gene_GOs.txt", 
            sep="\t",
            col.names = F, 
            row.names = F,
            quote = FALSE,
)
write.table(KO_df, file = "LOC_gene_KOs.txt", 
            sep="\t",
            col.names = F, 
            row.names = F,
            quote = FALSE
)
```



Reading in lists for mapping
```{r}

#ReadMapping of GO/KEGG
GO_db <- readMappings("clusterprofiler/LOC_gene_GOs.txt",
                      sep = "\t",
                      IDsep = ","
                      )

KEGG_db <- readMappings("clusterprofiler/LOC_gene_KOs.txt",
                        sep="\t", 
                        IDsep = ",")

```



Making GO and KO lists (Gene to GO, GO to term)
- Remove commas from values in a column and place separated values into new rows
  - https://stackoverflow.com/questions/17583926/r-remove-commas-from-values-in-a-column-and-place-separated-values-into-new-ro?rq=3
```{r}
GOs <- data.frame(cbind(
  unlist(lapply(strsplit(GO_df$GOID,","), cbind))
)) %>% unique() 
names(GOs)[1] <- "GOID"
KOs <- data.frame(cbind(
  unlist(lapply(strsplit(KO_df$KOs,","), cbind))
)) %>% unique() 
names(KOs)[1] <- "KOID"
```



Mapping genes to GO/KO terms
```{r}
#GOs
GOpath<-data.frame()
for(i in names(GO_db)){
  y =GO_db[[i]]
  ln<-length(GO_db[[i]])
  x=data.frame(ID=rep(i,ln),y)
  GOpath<-rbind(GOpath,x)
}
colnames(GOpath)<-c("ID","GOID")

head(GOpath)
GOpath[60:80,]
GO.Gene2Term <- GOpath


#KOs
KEGGpath<-data.frame()
for(i in names(KEGG_db)){
  y =KEGG_db[[i]]
  ln<-length(KEGG_db[[i]])
  x=data.frame(ID=rep(i,ln),y)
  KEGGpath<-rbind(KEGGpath,x)
}
KEGGpath$y<-sub("ko:","",KEGGpath$y)
colnames(KEGGpath)<-c("ID","KEGG")

head(KEGGpath)
KEGGpath[60:80,]

#saving these so don't have to rerun all this every time!
write.csv(GO.Gene2Term, "clusterprofiler/GO.Gene2Term.csv",
          col.names = T,
          row.names = F,
          sep = "\t")
write.csv(KEGGpath, "clusterprofiler/KEGGpath.csv",
          col.names = T,
          row.names = F,
          sep = "\t")
```



For GOs, add annotations to each GO to create Term2Name
```{r}
library(GO.db)
goterms <- Term(GOTERM)
goontology <- Ontology(GOTERM)
GO.Term2Name <- data.frame("GOID" = names(goterms),
                           "TERM" = goterms, 
                           "Ontology" = goontology )

#subsetting for BP, MF, CC GO terms
GO.Term2Name.CC = subset(GO.Term2Name, Ontology == "CC")
GO.Term2Name.MF = subset(GO.Term2Name, Ontology == "MF")
GO.Term2Name.BP = subset(GO.Term2Name, Ontology == "BP")

#and delete rows with word ribosom (ribosome AND ribosomal)
#https://stackoverflow.com/questions/22249702/delete-rows-containing-specific-strings-in-r
GO.Term2Name.CC = GO.Term2Name.CC[!grepl("ribosom", GO.Term2Name.CC$TERM),]
GO.Term2Name.MF = GO.Term2Name.MF[!grepl("ribosom", GO.Term2Name.MF$TERM),]
GO.Term2Name.BP = GO.Term2Name.BP[!grepl("ribosom", GO.Term2Name.BP$TERM),]
```



Also must make Gene2Term which connects genes to GO terms/descriptions
```{r}
#merge LOC gene IDs, GO term, and descriptions
#https://stackoverflow.com/questions/70878527/r-how-to-subset-a-dataframe-based-on-another-dataframe
GO.Gene2Term.CC = merge(GOpath, GO.Term2Name.CC, by = c("GOID"))
GO.Gene2Term.MF = semi_join(GOpath, GO.Term2Name.MF, by = c("GOID"))
GO.Gene2Term.BP = semi_join(GOpath, GO.Term2Name.BP, by = c("GOID"))

#swap columns
df = GO.Gene2Term.CC
df = df[, c("GOID", "ID")]
GO.Gene2Term.CC = df

df = GO.Gene2Term.MF
df = df[, c("GOID", "ID")]
GO.Gene2Term.MF = df

df = GO.Gene2Term.BP
df = df[, c("GOID", "ID")]
GO.Gene2Term.BP = df
```




With all the setup, can now make figures!


################################
ORA with GO terms via enricher()
################################
```{r}
#examining top white body genes 

#read in truly unique hematopoietic sequences 
Truly.Hemato.Unique <- file.choose() 
#choose Unique_hemato_degs.txt - might need to pull from venny still
#pulled from venny, choose this file instead: Postvenny_hemato_degs copy.txt
test <- fread(Truly.Hemato.Unique,
              header = F) 
Truly.Hemato.Unique <- test$V1


#establish universe of unique potential genes
universe <- GO.Gene2Term$GENE %>%
  unique()
gene <- Truly.Hemato.Unique 

BP.enrich <- enricher(gene = gene,
                   universe = universe,
                   pvalueCutoff = .05,
                   pAdjustMethod = "BH",
                   minGSSize = 10,
                   maxGSSize = 500, #(length of gene)
                   qvalueCutoff = .02,
                   TERM2GENE = GO.Gene2Term.BP,
                   TERM2NAME = GO.Term2Name.BP
)
CC.enrich <- enricher(gene = gene,
                   universe = universe,
                   pvalueCutoff = .05,
                   pAdjustMethod = "BH",
                   minGSSize = 10,
                   maxGSSize = 500, #(length of gene)
                   qvalueCutoff = .02,
                   TERM2GENE = GO.Gene2Term.CC,
                   TERM2NAME = GO.Term2Name.CC
)
MF.enrich <- enricher(gene = gene,
                   universe = universe,
                   pvalueCutoff = .05,
                   pAdjustMethod = "BH",
                   minGSSize = 10,
                   maxGSSize = 500, #(length of gene)
                   qvalueCutoff = .02,
                   TERM2GENE = GO.Gene2Term.MF,
                   TERM2NAME = GO.Term2Name.MF
)

BP.dotplot <- dotplot(BP.enrich, showCategory = 20)
CC.dotplot <- dotplot(CC.enrich, showCategory = 20)
MF.dotplot <- dotplot(MF.enrich, showCategory = 20)
aplot::plot_list(BP.dotplot,
                   CC.dotplot,
                   MF.dotplot
                   #labels=LETTERS[1:3]
                 )
plot_grid(BP.dotplot,
          CC.dotplot,
          MF.dotplot,
          ncol = 1, align = "v")


cnetplot(BP.enrich, 
         #categorySize="pvalue", 
         #foldChange=$log2FoldChange
         node_label="category"
         )
         
cnetplot(CC.enrich, 
         #categorySize="pvalue"#, 
         #foldChange=$log2FoldChange
         )
cnetplot(MF.enrich, 
         #categorySize="pvalue"#, 
         #foldChange=$log2FoldChange
         )

#heatplot
heatplot(BP.enrich, showCategory = 10)
heatplot(CC.enrich, showCategory = 10)
heatplot(MF.enrich, showCategory = 10)


# calculate pairwise similarities of the enriched terms using Jaccard’s similarity index
BP.enrich.Jaccard <- pairwise_termsim(BP.enrich)
BP.treeplot <- treeplot(BP.enrich.Jaccard, showCategory = 20)
CC.enrich.Jaccard <- pairwise_termsim(CC.enrich) 
CC.treeplot <- treeplot(CC.enrich.Jaccard, showCategory = 20)
MF.enrich.Jaccard <- pairwise_termsim(MF.enrich)
MF.treeplot <- treeplot(MF.enrich.Jaccard, showCategory = 20)

aplot::plot_list(BP.treeplot, 
                 CC.treeplot,
                 MF.treeplot
                 #tag_levels='A'
                 )
p3 <- plot_grid(BP.treeplot, 
                 CC.treeplot,
                 MF.treeplot,
          ncol = 1, align = "v")

# Enrichment map 
BP.emapplot <- emapplot(BP.enrich.Jaccard, layout = "kk", 
                        showCategory = 20, 
                        node_label = 'group',
                        nCluster = 5
)
CC.emapplot <- emapplot(CC.enrich.Jaccard, layout = "kk", 
                        showCategory = 20, 
                        node_label = 'group',
                        nCluster = 5
)
MF.emapplot <- emapplot(MF.enrich.Jaccard, layout = "kk", 
                        showCategory = 20, 
                        node_label = 'group',
                        nCluster = 5
)
aplot::plot_list(BP.emapplot,
                   CC.emapplot,
                   MF.emapplot
                   #labels=LETTERS[1:3]
                 )
plot_grid(BP.emapplot,
                   CC.emapplot,
                   MF.emapplot,
          ncol = 1, align = "v")

#cnetplot again
BP.cnetplot <- cnetplot(BP.enrich.Jaccard, 
         #categorySize="pvalue", 
         #circular = TRUE, 
         #colorEdge = TRUE
         )
         
CC.cnetplot <- cnetplot(CC.enrich.Jaccard, 
         categorySize="pvalue",
         circular = TRUE, colorEdge = TRUE
)
MF.cnetplot <- cnetplot(MF.enrich.Jaccard, 
         categorySize="pvalue",
         circular = TRUE, colorEdge = TRUE
         )





outpathcount = "clusterprofiler/"
write.csv(BP.enrich, file=paste(outpathcount,
                                "Top.WB.DEGs.BP.enrich.csv",
                                sep = ""), 
          row.names = T, col.names = T, quote=FALSE) 
write.csv(CC.enrich, file=paste(outpathcount,
                                "Top.WB.DEGs.CC.enrich.csv",
                                sep = ""), 
          row.names = T, col.names = T, quote=FALSE) 
write.csv(MF.enrich, file=paste(outpathcount,
                                "Top.WB.DEGs.MF.enrich.csv",
                                sep = ""), 
          row.names = T, col.names = T, quote=FALSE) 
write.csv(BP.enrich.Jaccard, file=paste(outpathcount,
                                "Top.WB.DEGs.BP.enrich.Jaccard.csv",
                                sep = ""), 
          row.names = T, col.names = T, quote=FALSE) 
write.csv(CC.enrich.Jaccard, file=paste(outpathcount,
                                "Top.WB.DEGs.CC.enrich.Jaccard.csv",
                                sep = ""), 
          row.names = T, col.names = T, quote=FALSE) 
write.csv(MF.enrich.Jaccard, file=paste(outpathcount,
                                "Top.WB.DEGs.MF.enrich.Jaccard.csv",
                                sep = ""), 
          row.names = T, col.names = T, quote=FALSE) 




```




##############################
ORA with KO terms via enKEGG()
##############################

Thsi tutorial super helpful: https://orlanc.github.io/cp_enrichment/R_enrichment/kegg_pathway_enrichment.html

```{r}

id <- Truly.Hemato.Unique

id_ko<-KEGGpath[which(KEGGpath$ID%in%id),]
head(id_ko, n = 10)

#Delete those IDs that have non-Knumber associated
ko<-id_ko[id_ko$KEGG!="-",]
head(ko)

#remove duplicate rows if any
#https://www.datanovia.com/en/lessons/identify-and-remove-duplicate-data-in-r/
ko <- ko %>% distinct(ID, KEGG, .keep_all = TRUE)


#Perform the KEGG pathway enrichment analysis
#Use Knumber from data frame ko. 
#In this step, you might get an error; if it happens, go back to Install and load packages.

knum<-ko$KEGG
enKEGG <-enrichKEGG(knum, 
                     organism = 'ko', 
                     minGSSize = 10,
                     maxGSSize = 500,
                     keyType = "kegg", 
                     pvalueCutoff = .05,
                     pAdjustMethod = "BH", 
                     qvalueCutoff = .02
)

write.csv(enKEGG, file=paste(outpathcount,
                                 "Top.WB.enKEGG.csv",
                                 sep = ""), 
            row.names=TRUE, col.names=T, quote=FALSE) 

#getting errors when trying to run emapplot, likely because of duplicate terms (ie cell cycle in jaccard df)
#try to filter out w clusterprofiler's function
# Filter out duplicates so each Description only appears once
enKEGG <- clusterProfiler::filter(enKEGG, !duplicated(Description))

enKEGG.dotplot <- dotplot(enKEGG, showCategory = 20)

enKEGG.Jaccard <- pairwise_termsim(enKEGG)
write.csv(enKEGG.Jaccard, file=paste(outpathcount,
                                "Top.WB.DEGs.enKEGG.Jaccard.csv",
                                sep = ""), 
          row.names = T, col.names = T, quote=FALSE) 
enKEGG.treeplot <- treeplot(enKEGG.Jaccard, showCategory=20)


enKEGG.emapplot <- emapplot(enKEGG.Jaccard,
                            showCategory = 20,
                            #min_edge = 0,
                            layout = "kk",
                            node_label = 'group',
                            nCluster = 5
                            #group = TRUE
                            #cex.params = list(category_label = 0)
                            )
```


#############################
Saving combined GO/KEGG plots
#############################
```{r}
p1 <- plot_grid(BP.dotplot,CC.dotplot,MF.dotplot,enKEGG.dotplot,
          ncol=1,
          align="v")
# Save it as an SVG
ggsave(filename = "GO_KEGG_dotplot.svg", 
       plot = p1, 
       device = "svg",      # Forces SVG engine
       width = 6,          # Broad width prevents label clipping
       height = 33,         # Broad height helps spread out the network
       units = "in",        # Dimensions in inches
       bg = "white")        # Keeps the background solid white



p2 <- aplot::plot_list(BP.emapplot,CC.emapplot,MF.emapplot,enKEGG.emapplot,
          ncol=1
          #align="v"
          )
# Add 1-inch margins around the entire plot to fit the ellipses
p3 <- enKEGG.emapplot + theme(plot.margin = margin(.5, .5, .5, .5, "in"))
# Allow shapes to draw completely outside the axes bounding box
p4 <- p3 + coord_cartesian(clip = "off")
# Save it as an SVG

#might have to save plots individually
ggsave(filename = "GO_KEGG_emapplot.svg", 
       plot = p4, 
       device = "svg",      # Forces SVG engine
       width =10,          # Broad width prevents label clipping
       height =6,         # Broad height helps spread out the network
       units = "in",        # Dimensions in inches
       bg = "white",         # Keeps the background solid white
       scale = 0.9
)





p3 <- plot_grid(BP.treeplot,CC.treeplot,MF.treeplot,enKEGG.treeplot,
          ncol=1,
          align="v")
# Save it as an SVG
ggsave(filename = "GO_KEGG_treeplot.svg", 
       plot = p3, 
       device = "svg",      # Forces SVG engine
       width = 18,          # Broad width prevents label clipping
       height = 33,         # Broad height helps spread out the network
       units = "in",        # Dimensions in inches
       bg = "white")        # Keeps the background solid white





pAll <- plot_grid(
          BP.dotplot,
          BP.treeplot,
          BP.emapplot,
          
          CC.dotplot,
          CC.treeplot,
          CC.emapplot,
          
          MF.dotplot,
          MF.treeplot,
          MF.emapplot,
          
          enKEGG.dotplot,
          enKEGG.treeplot,
          enKEGG.emapplot,
          
          ncol = 3, 
          nrow = 4,
          align = "v")

ggsave(filename = "GO_KEGG_allplots.svg", 
       plot = pAll, 
       device = "svg",      # Forces SVG engine
       width = 30,          # Broad width prevents label clipping
       height = 40,         # Broad height helps spread out the network
       units = "in",        # Dimensions in inches
       bg = "white")        # Keeps the background solid white



```




####################################################
Gene Set Enrichment Analysis (GSEA) for M/F WB genes
####################################################
For this, we need:
  - a list of genes were interested in 
  - a ranking of those genes (in this case we will take L2FC)
  
Previously, got error when running this, but since the analysis was re-leveled on F.WBs up top when making the results, shouldn't be an issue
```{r}

#getting just shrunken LFC values
resLFC <- res.df[c("log2FoldChange")]

#gene list setup
PreGeneList <- tibble::rownames_to_column(resLFC.LFC, "GENE") %>% as.data.frame()
geneList = PreGeneList[,2]
names(geneList) = as.character(PreGeneList[,1])
geneList = sort(geneList, decreasing = TRUE)

#swap columns maybe? -> fixed this up top
#df = GO.Gene2Term
#df = df[, c("GOID", "ID")]
#GO.Gene2Term = df

gsea <-GSEA(
  geneList,
  exponent = 1,
  minGSSize = 10,
  maxGSSize = 500,
  eps = 0,
  pvalueCutoff = 1,
  pAdjustMethod = "none",
  gson = NULL,
  TERM2GENE = GO.Gene2Term,
  TERM2NAME = GO.Term2Name.MF,
  verbose = TRUE,
  seed = FALSE,
  by = "fgsea"
)
#don't need to do a BH or low p adj cause these genes are already sig. 
#Just want to see how FC affects representative terms

write.csv(gsea, file="GSEA_res_CC.csv", 
          #sep="\t",
          #col.names = F, 
          #row.names = F,
          quote = FALSE) 

#Dotplot
dotplot(gsea, showCategory=20, split=".sign") + facet_grid(.~.sign)

#Encrichment Map:Enrichment map organizes enriched terms into a 
#network with edges connecting overlapping gene sets. In this way, 
#mutually overlapping gene sets are tend to cluster together, 
#making it easy to identify functional modules.
#https://github.com/YuLab-SMU/enrichplot/issues/79
gsea_emap <- enrichplot::pairwise_termsim(gsea)#, semData = d)
emapplot(gsea_emap)


# categorySize can be either 'pvalue' or 'geneNum'
#this is th best one cayse we can change the parameter to LFC which is what is interesting
#we don't need to do a BH p adj. cause these genes are already sig. Just want to see how FC affects representative terms

library(igraph)
p1 <- cnetplot(gsea, 
               showCategory = 25,
               foldChange = geneList,
               layout = igraph::layout_nicely)

# Add your custom color scale (e.g., standard Blue-White-Red gradient)
p2 <- p1 + scale_color_gradient2(name = "Log 2 Fold Change", 
                          low = "deeppink2", 
                          mid = "gray", 
                          high = "deepskyblue", 
                          midpoint = 0)
p2

# Save it as an SVG
ggsave(filename = "cnetplot_gsea.svg", 
       plot = p2, 
       device = "svg",      # Forces SVG engine
       width = 15,          # Broad width prevents label clipping
       height = 13,         # Broad height helps spread out the network
       units = "in",        # Dimensions in inches
       bg = "white")        # Keeps the background solid white
```


























